---
title: "dplyr and tidyr"
subtitle: "(MPA 6020)"
author: "Ani Ruhil"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "rladies-fonts"]
    seal: true
    self_contained: false
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightlines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(fig.align = 'center', echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, tidy = TRUE, tidy.opts = list(width.cutoff = 90), fig.align = "center", fig.width = 8, fig.height = 6, out.width = "60%", cache = TRUE) 
```

## Agenda 
- using `dplyr` 
- using `tidyr` 

---

class: inverse, middle, center 

<center><img src = "./images/hex-dplyr.png", width = 200px></center>

---

"[dplyr](http://dplyr.tidyverse.org/) is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges" 

The core functions are: 

- mutate() ... to add new variables that are functions of existing variables 
- transmute() ... to add new variables and drop the rest 
- select() ... to pick variables (based on their names) to work on 
- filter() ... to pick cases (based on their values) to work on 
- summarise() .. to calculate quantities of interest (rather than add these quantities to the existing data frame) and save these as a standalone object 
- arrange() ... order the rows based on out preferences 

In order to understand how these commands work we rely on a simple data frame -- all flights originating and departing from Columbus (Ohio) January through September of 2017. Let  us load the data and `dplyr`. 

```{r cmhflights}
load("./data/cmhflights_01092017.RData")

library(dplyr)
```

You should see 110 columns, the last being `X110`, an empty column so we drop it. 

```{r dropx110}
cmhflights$X110 = NULL
```

---
## select 

Say I only want the first five columns. 

```{r select01}
my.df = cmhflights %>% 
  select(Year:DayOfWeek)
names(my.df)
```

What if the columns were not sequentially located? 

```{r select02}
my.df = cmhflights %>% 
  select(Year, FlightDate:UniqueCarrier, TailNum)
names(my.df)
```

Could we use column numbers instead? 

```{r select03}
my.df = cmhflights %>% 
  select(c(1, 3, 5, 7))
names(my.df)
```

---

You can also select columns in other ways:

```{r select04}
my.df = cmhflights %>% 
  select(contains("Carrier"))
names(my.df)

my.df = cmhflights %>% 
  select(starts_with("De"))
names(my.df)

my.df = cmhflights %>% 
  select(ends_with("Num"))
names(my.df)
```

---
## filter

We used `subset()` to select particular cases for analysis. With `dplyr`, the `filter()` command serves the same purpose. Say we only want records from January, i.e., `Month == 1`. 

```{r filter01}
my.df = cmhflights %>% 
  filter(Month == 1)
table(my.df$Month)
```

What about only American Airline flights in January? 

```{r filter02}
my.df = cmhflights %>% 
  filter(Month == 1 & UniqueCarrier == "AA")
table(my.df$Month, my.df$UniqueCarrier)
```

What about United Airlines flights in January to CMH (Columbus, OH) to any destination? 

```{r filter03}
my.df = cmhflights %>% 
  filter(Month == 1 & UniqueCarrier == "UA" & Dest == "CMH") 
```

---

What if I wanted a more complicated filter, say, flights in January or February to CMH or ORD?

```{r filter04}
my.df = cmhflights %>% 
  filter(Month %in% c(1, 2) & UniqueCarrier == "UA" & Dest %in% c("CMH", "ORD")) 
table(my.df$Month)
table(my.df$UniqueCarrier)
table(my.df$Dest)
```

---
Say I wanted to arrange the resulting data frame by ascending order of departure delays. How might I do that? 

## arrange

```{r arrange01}
my.df2 = my.df %>% 
  arrange(DepDelayMinutes)
```

And now in descending order of delays. 

```{r arrange02}
my.df2 = my.df %>% 
  arrange(-DepDelayMinutes)
```

We could tweak this further, perhaps saying sort by departure delays to CMH, and then to ORD. 

```{r arrange03}
my.df2 = my.df %>% 
  arrange(Dest, -DepDelayMinutes)
```

---

So far, we have seen each function in isolation. Now we streamline things a bit so that we only end up with the columns we want. That is, we can string together these functions to have a slim data frame of only those columns we selected, filtered as we want, and arranged as we want. 


```{r stringing01}
my.df3 = cmhflights %>% 
  select(Month, UniqueCarrier, Dest, DepDelayMinutes) %>% 
  filter(Month %in% c(1, 2) & UniqueCarrier == "UA" & Dest %in% c("CMH", "ORD")) %>% 
  arrange(Month, Dest, -DepDelayMinutes)
```

Here, the end result is a data frame arranged by Month, then within Month by Destination, and then finally by descending order of flight delays. This is the beauty of `dplyr`, allowing us to chain together various functions to get what we want. 

---
## group_by

Let us calculate the Mean and Median delay. To do this, we will lean on the `group_by()` function to calculate median and mean delays by Month and Destination. While we are at it, let us see the standard deviation as well. 

```{r gby01}
tab.01 = cmhflights %>% 
  select(Month, UniqueCarrier, Dest, DepDelayMinutes) %>% 
  filter(Month %in% c(1, 2) & UniqueCarrier == "UA" & Dest %in% c("CMH", "ORD")) %>% 
  group_by(Month, Dest) %>% 
  summarise(Median.Delay = median(DepDelayMinutes, na.rm = TRUE), 
            Mean.Delay = mean(DepDelayMinutes, na.rm = TRUE),
            SD.Delay = sd(DepDelayMinutes, na.rm = TRUE)) %>% 
  arrange(Month, Dest)
```

What if wanted to know how many airports can you fly to from Columbus, and how many flights there were to each in July? 

```{r gby02}
tab.02 = cmhflights %>% 
  select(Month, UniqueCarrier, Dest, DepDelayMinutes, Origin) %>% 
  filter(Month %in% c(7) & Origin %in% c("CMH")) %>% 
  group_by(Dest) %>% 
  summarise(No.of.Flights = n()) 
```

---

How many distinct airlines fly out of Columbus? 

```{r gby03}
tab.03 = cmhflights %>% 
  select(Month, UniqueCarrier, Dest, DepDelayMinutes, Origin) %>% 
  filter(Origin %in% c("CMH")) %>% 
  group_by(UniqueCarrier) %>% 
  summarise(No.of.Airlines = n_distinct(UniqueCarrier)) 
```

And how many flights did each airline operate? 

```{r gby04}
tab.04 = cmhflights %>% 
  select(Month, UniqueCarrier, Dest, DepDelayMinutes, Origin) %>% 
  filter(Origin %in% c("CMH")) %>% 
  group_by(UniqueCarrier) %>% 
  summarise(No.of.Flights = n()) 
```

Of all of these airlines, which one has the highest mean departure delay? 

```{r gby05}
tab.05 = cmhflights %>% 
  select(Month, UniqueCarrier, Dest, DepDelayMinutes, Origin) %>% 
  filter(Origin %in% c("CMH")) %>% 
  group_by(UniqueCarrier) %>% 
  summarise(Median.Delay = mean(DepDelayMinutes, na.rm = TRUE)) 
```

---

## mutate
Thus far we have been creating a summary data frame containing whatever we calculated -- the frequencies, averages, and so  on. Some times, however, you may want these calculations to be be appended to the original data frame and stored as new columns. This can be done with `mutate()`, as shown below. In this particular example, I am calculating median departure delay by airline (for each airport it departs from) and then including this as a column called `Median.Delay` in cmhflights. 

```{r mut01}
cmhflights = cmhflights %>% 
  group_by(UniqueCarrier, Origin) %>% 
  mutate(Median.Delay = mean(DepDelayMinutes, na.rm = TRUE)) 
```

Note the 110th column. Now, if we wanted to add this column but drop all other columns not being used in the calculation, we could do that via `transmute()`. Be careful with this one though. 

```{r mut02}
mut.df = cmhflights

mut.df2 = mut.df %>% 
  group_by(UniqueCarrier, Dest) %>% 
  transmute(Median.Arr.Delay = mean(ArrDelayMinutes, na.rm = TRUE)) 

names(mut.df2)
```

---

class: inverse, middle, center

### Calculations with dplyr 

---

I want a simple frequency table of flights by carrier. 

```{r calc01}
tab.01 = cmhflights %>% 
  count(Carrier, sort =  TRUE)
```

By carrier and destination? 

```{r calc02}
tab.02 = cmhflights %>% 
  count(Carrier, Dest, sort =  TRUE)
```

How many unique aircraft do we have?

```{r calc03}
cmhflights %>% 
  summarise(unique.planes = n_distinct(TailNum)) 
```

---

How many flights do we see per unique aircraft? 

```{r calc04}
tab.03 = cmhflights %>% 
  group_by(TailNum) %>% 
  summarise(Number = n()) 
```

If you wanted options in terms of carriers when flying out of Columbus, which destination would give you maximal options? 

```{r calc05}
tab.04 = cmhflights %>% 
  filter(Origin == "CMH") %>% 
  group_by(Dest) %>% 
  summarise(unique.carriers = n_distinct(Carrier)) %>% 
  arrange(-unique.carriers)
```

How many flights to each destination, from Columbus, on February 15?

```{r calc06}
tab.05 = cmhflights %>% 
  filter(Month == 1 & DayofMonth == 1 & Origin == "CMH") %>% 
  group_by(Dest) %>% 
  summarise(n.flts = n()) %>% 
  arrange(-n.flts)
```


What if we want the preceding by airline?

```{r calc07}
tab.06 = cmhflights %>% 
  filter(Month == 1 & DayofMonth == 1 & Origin == "CMH") %>% 
  group_by(Dest, Carrier) %>% 
  summarise(n.flts = n()) %>% 
  arrange(-n.flts)
```

---

class: inverse, middle, center

### unnest

---

# unnest 
Every now and then the data you have to work with includes some list-columns -- columns that are like mini-data frames themselves. Suddenly R functions that would otherwise work will not execute as desired. The `unnest` function from the `tidyverse` unpacks these list-columns into a regular tibble. First a small data frame that shows a list-column. 

```{r un1}
library(tidyverse)
bad.df = data_frame(x = 1:3, y = c("a", "d,e,f", "g,h"))
```

If I want to clean up list-column `y`, I can do that as shown below. 

```{r un2}
good.df = bad.df %>% 
  transform(y = strsplit(y, ",")) %>% 
  unnest(y)
```

Now see a slightly more complicated list-column, with `y` made of named rows .... a and b, a with 1 value and b with 2 values. 

```{r un3}
badder.df = data_frame(x = 1:2, y = list(a = 1, b = 3:4))
```

To clean this up we could do: 

```{r un4}
gooder.df = badder.df %>% 
  unnest(., .id = "name")
```

---

class: inverse, middle, center 

<center><img src = "./images/hex-tidyr.png", width = 200px></center>

---

Most packages you use in R call for data to be `tidy`. [What are tidy data](http://vita.had.co.nz/papers/tidy-data.pdf)? 

- Each variable forms a column. 
- Each observation forms a row. 
- Each type of observational unit forms a table. 

Here are some untidy data 

(a) The Pew survey data 

```{r untidy01}
pew = read.delim(
  file = "http://stat405.had.co.nz/data/pew.txt",
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = F
)

head(pew)
```

---

(b) Here are some more examples 

```{r untidy02a}
library(DSR)
table2
```

---

```{r untidy02b}
table3
```

```{r untidy02c}
table4
```

```{r untidy02d}
table5
```

---

What would a tidy version of the examples in (b) look like? 

```{r untidy03}
table1
```

Here, we have 

- each country as a row per year 
- each column has a single variable 
- each cell has a single value 

The `tidyr` package is designed to make messy data tidy in order to proceed with analysis. It has four primary functions that we will see in turn. 

---

## `separate` 
You will run into variables that you want to break into smaller pieces. For instance, take `table3` where `rate` is a ratio of two variables, `cases` and `population`, each of which should be in their own columns. How can we achieve that? 

```{r tidy01}
library(tidyr)
s.table3 = table3 %>% 
  separate(col = rate, into = c("cases", "population"), sep = "/", 
           remove = TRUE, convert = TRUE) 
```

## `unite`

The `unite` command does the exact opposite of separate, combining columns 

```{r tidy02}
u.table3 = s.table3 %>% 
  unite("rate", cases, population, sep = ":", remove = FALSE)
```

Here you specify the name of  the new column first, and then list the columns to be combined, specify the separator, and whether the original columns should be removed or not. 

---

## `spread` 

Revisit `table2` 

```{r tidy03}
table2
```

---

I want to flip this into a format where the data frame has two columns per country per year -- `cases` and `population`. How can I do this? 

```{r tidy04}
sp.table2 = table2 %>% 
  spread(key, value)
sp.table2
```

Notice that the  `key` column had two unique values -- cases and population -- so two new columns were created. Notice also that tidyr knew to populate the value of each of the new cells with what it finds in the `value` column. The result is what we call data in a wide format. 

---

## `gather` 

`gather` does the opposite of `spread`, converting wide data into what we call the long form. Revisit `table4`

```{r tidy05}
table4
```

Note that each country has two columns, one per  year. This is untidy. So we'll flip this tibble as follows:  

```{r tidy06}
g.table4 = table4 %>% 
  gather("year", "cases", 2:3)
```

Notice that you had to specify the columns you wanted to gather and we did this via `2:3`. If you forget to specify columns it creates something else altogether so be careful. 

---

# WHO Example  

[At this website](http://www.who.int/tb/country/data/download/en/) you can find data provided by countries to WHO and estimates of TB burden generated by WHO for the Global Tuberculosis Report. I want you to gather the following data (all are in the csv format): 

- [Download the data dictionary](https://extranet.who.int/tme/generateCSV.asp?ds=dictionary) 
- [Download case notifications](https://extranet.who.int/tme/generateCSV.asp?ds=notifications)  
- [Download WHO TB burden estimates](https://extranet.who.int/tme/generateCSV.asp?ds=estimates) 

Read the data dictionary carefully. Then read-in the case notifications data. 

The database has too many variables (175 or so) and so I'll have you use just a few for this practice task. 

```{r pt01}
library(readr)
tb = read_csv("../data/TB_notifications_2018-02-07.csv")
```

---

Let us use `dplyr` to select and filter the tb database so that it meets the following criteria: (i) observations from 2012 or earlier years are kept, and (ii) keep only columns 1 through 6 and 25 (new_sp_m04) through 44 (new_sp_fu).  

```{r pt02}
library(dplyr)
tb2 = tb %>% 
  select(c(1:6, 25:44)) %>% 
  filter(year <= 2012)
names(tb2)
head(tb2, 3)
```

---

Now, `new_sp_xx` refer to new pulmonary smear-positive cases. The `m` refers to males and `f` refers to females. Finally, `04` is the 0-4 age group, `514` is the 5-14 age group, and so on. `new_sp_mu` is males with an unknown age-group and `new_sp_fu` is females with an unknown age-group. This is the messy part of the data. Ideally we'd have a column for gender, flagged as `Male, Female, Unknown`, and a separate column for age-groups flagged `0-4, 5-14, 0-14, and so on`. Time to tidy these data. 

Since the data are in the wide format we can use gather to flip it into the long format. 

```{r pt03}
tb3 = tb2 %>% 
  gather("group", "cases", 7:26) 

head(tb3)
```

---

Now we split the `new_sp_xx` values into three columns 

```{r pt04}
tb4 = tb3 %>% 
  separate(col = group, into = c("new", "sp", "sexage"), sep = "_") 

head(tb4)
```

---

and then split `sexage` into two columns by specifying that the split should occur after the first value (which in our case will be after `m` or `f`). 

```{r pt05}
tb5 = tb4 %>% 
  separate(col = sexage, into = c("sex", "agegroup"), sep = 1) 

head(tb5)
```

---

The beauty of dplyr and tidyr is the ability to combine all these steps into a single piped command! 

```{r pt06}
tb.df = tb %>% 
  select(c(1:6, 25:44)) %>% 
  filter(year <= 2012)  %>% 
  gather("group", "cases", 7:26) %>% 
  separate(col = group, into = c("new", "sp", "sexage"), sep = "_") %>% 
  separate(col = sexage, into = c("sex", "agegroup"), sep = 1) %>% 
  select(c(1:6, 9:11)) 

head(tb.df)
```

---

We can clean-up the codes for easier interpretation. 

```{r pt07} 
tb.df$sex = ifelse(tb.df$sex == "m", "Males", "Females") 

tb.df$agegroup = ifelse(tb.df$agegroup == "04", "0-4", 
                        ifelse(tb.df$agegroup == "014", "0-14", 
                        ifelse(tb.df$agegroup == "1524", "15-24",
                        ifelse(tb.df$agegroup == "2534", "25-34",
                        ifelse(tb.df$agegroup == "3544", "35-44",
                        ifelse(tb.df$agegroup == "4554", "45-54",
                        ifelse(tb.df$agegroup == "514", "5-14",
                        ifelse(tb.df$agegroup == "5564", "55-64",
                        ifelse(tb.df$agegroup == "65", "65+", "Uknown")))))))))

head(tb.df)
```

---
How many cases per country in the 1980-1990 period? 

```{r pt08}
tbtab.01 = tb.df %>% 
  filter(year >= 1980 | year <= 1990) %>% 
  group_by(country) %>% 
  summarise(Ncases = sum(cases, na.rm = TRUE)) %>% 
  arrange(-Ncases)

head(tbtab.01)
```

---

What about by sex and age-group?

```{r pt09}
tbtab.02 = tb.df %>% 
  filter(year >= 1980 | year <= 1990) %>% 
  group_by(country,  sex, agegroup) %>% 
  summarise(Ncases = sum(cases, na.rm = TRUE)) %>% 
  arrange(-Ncases)

head(tbtab.02)
```

This is just one example of how `dplyr` and `tidyr` -- two packages from the `tidyverse` -- can be used to deal with what seems to be quite messy data. The more you use it, the easier it becomes to figure out solutions to even very complicated messy data setups. 

---

class: inverse, middle, center 

<center><img src = "./images/hex-forcats.png" width = 200px></center>

---

Say I have the `mtcars` data and want to see the distribution of the number of cylinders a car has

```{r fc01}
data(mtcars)
table(mtcars$cyl)
```

`cyl` is stored as a numeric so I can flip this into a factor via 

```{r fc02}
library(forcats)
mtcars$f.cyl = factor(mtcars$cyl)
mtcars$f.cyl = fct_recode(mtcars$f.cyl, Four = '4', Six = '6', Eight = '8')
table(mtcars$f.cyl)
```

---

If I plot `mpg` by the car, I get 

```{r fc03, fig.width = 6, fig.height = 4, out.width="40%"}
mtcars$model = row.names(mtcars)

library(ggplot2)
ggplot(mtcars, aes(x = mpg, y = model)) + geom_point() + labs(x = "Miles per gallon", y = "Car Model")
```

This isn't very helpful since the models are not arranged in ascending/descending order of miles per gallon. That can be easily fixed:

---

.pull-left[

If I  want the plot to be in descending order of `mpg`, 

```{r fc04, fig.width = 6, fig.height = 4, out.width="80%"}
ggplot(mtcars, aes(x = mpg, fct_reorder(f = model, x = mpg))) + geom_point() + labs(x = "Miles per gallon", y = "Car Model")
```
]

.pull-right[

If I  want the plot to be in descending order of `mpg`, 

```{r fc05, fig.width = 6, fig.height = 4, out.width="80%"}
ggplot(mtcars, aes(x = mpg, fct_reorder(f = model, x = -mpg))) + geom_point() + labs(x = "Miles per gallon", y = "Car Model")
```
]

---

I can do this with bar-plots too. 

```{r fc06}
ggplot(mtcars, aes(f.cyl)) + geom_bar(aes(fill = f.cyl)) + theme(legend.position = "none") + labs(x = "Number of cylinders", y = "Frequency")
```

---

.pull-left[
The bars ordered by descending frequency

```{r fc07a}
ggplot(mtcars, aes(fct_infreq(f.cyl))) + geom_bar(aes(fill = f.cyl)) + theme(legend.position = "none") + labs(x = "Number of cylinders", y = "Frequency")
```
]

.pull-right[
The bars ordered by ascending frequency
```{r fc07b}
ggplot(mtcars, aes(fct_rev(fct_infreq(f.cyl)))) + geom_bar(aes(fill = f.cyl)) + theme(legend.position = "none") + labs(x = "Number of cylinders", y = "Frequency")
```
]

Note the use of `fct_infreq()` for descending and then `fct_rev(fct_infreq())` for ascending. 
