---
title: "Working with dates and times"
subtitle: "(MPA 6020)"
author: "Ani Ruhil"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "rladies-fonts"]
    seal: true
    self_contained: false
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightlines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(fig.align = 'center', echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, tidy = TRUE, tidy.opts = list(width.cutoff = 90), fig.align = "center", fig.width = 8, fig.height = 6, out.width = "60%", cache = TRUE) 
```

## Agenda 

Understanding how to work with dates and times with 

- using `base R` 
- using `lubridate` 

---

Start by creating some data values.  

```{r dates00}
basedate = as.Date("1970-1-1")
today = Sys.Date()
tomorrow = as.Date("2017-12-18")
yesterday1 = "12/16/2017"
yesterday2 = "12-16-17"
yesterday3 = "12 16 17"
```

How did R read these? 

```{r dates01}
str(list(basedate, today, tomorrow, yesterday1, yesterday2, yesterday3))
```

Some are dates some are characters `(chr)` 

---

We can flip the characters into dates 

```{r dates02}
yesterday1d = as.Date(yesterday1, format = "%m/%d/%Y")
yesterday2d = as.Date(yesterday2, format = "%m-%d-%y")
yesterday3d = as.Date(yesterday3, format = "%m %d %y")
yesterday1d; yesterday2d; yesterday3d
```

---

There are special switches that indicate date components when you `format = ""`

```{r dates08}
df = data.frame(Code = c("%a", "%A", "%d", "%m", "%b", "%B", "%y", "%Y"), 
    Represents = c("Day spelled out (abbreviated)", "Day spelled out", "Day of the month", "Month", "Month (abbreviated)", 
        "Month (fully spelled out)", "Year (2-digits)", "Year (4-digits)"), 
    Example = c("Mon", "Monday", "16", "12", "Dec", "December", "17", "2017"))

df
```


---

So if you ever run into a date field that needs formatting, what we have covered should help you convert it into a proper date formatted variable.  

Once you convert them to the date format, you can extract other quantities. 

```{r dates03}
weekdays(yesterday1d)
months(today)
quarters(today)
```

---

Sometimes you will have a number representing the date, with the number representing the number of days since (if positive) or before (if negative) some date  of origin. R uses `1970-01-01` but other software packages may (and some do) use different days. Thus, for example, if the dates are listed as 

```{r dates04}
x = c(17651, -2345, 19760)
```

Then these numbers represent the following dates: 

```{r dates05}
x.dates = as.Date(x, origin = "1970-01-01")
x.dates
```

You can also convert dates into the numbers representing dates. 

```{r dates06}
x.julian = julian(x.dates, origin = as.POSIXct("1970-01-01", tz = "UCT"))
x.julian
```

[Julian?](http://aa.usno.navy.mil/data/docs/JulianDate.php) 

---

What if I wanted to create a sequence of dates, starting with Feb 8, 2018 but increasing by 1 day at a time or by every 5th day? 

```{r dates07}
start.date = as.Date(as.character("2018-02-08"), format = "%Y-%m-%d")
date.seq1 = seq(start.date, by = 1, length.out = 7)
date.seq1
date.seq5 = seq(start.date, by = 5, length.out = 3)
date.seq5
```

Note that `length.out =` is specifying how many you dates want to create. 

---

Now, what if I want to know the date 30 days from today? 19 days ago? 

```{r dates09}
date.30 = today + 30
date.19 = today - 19
```

How many days lapsed between two dates? 

```{r dates10}
date1 = as.Date("2017-04-28")
date2 = Sys.Date()
lapsed.time = date2 - date1; lapsed.time
```

Say I want to create a vector of dates that starts and ends on specific dates, and the step function is 1 day, a week, 4 months, etc. The step is indicated with the `by = ""` command 

```{r dates11}
my.dates1 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "day")
my.dates2 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "week")
my.dates3 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "month")
my.dates4 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "3 days")
my.dates5 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "2 weeks")
my.dates6 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "4 months")
my.dates7 = seq(from = as.Date("2017-12-17"), to = as.Date("2019-12-16"), by = "year")
my.dates8 = seq(from = as.Date("2017-12-17"), to = as.Date("2022-12-16"), by = "2 years")
```

---

class: inverse, middle, center 

<center><img src = "./images/hex-lubridate.png"></center>

---

This package makes working with dates and times quite easy because it does what base R does but in more intuitive ways and perhaps more flexibly. Let us start with some date fields.

```{r  lu01}
today1 = "20171217"
today2 = "2017-12-17"
today3 = "2017 December 17"
today4 = "20171217143241"
today5 = "2017 December 17 14:32:41"
today6 = "December 17 2017 14:32:41"
today7 = "17-Dec, 2017 14:32:41"
```

The formats are quite varied but `lubridate` deals with them quite seamlessly so long as you pay attention to the order -- is year first or last? What about month? day? Is time given in hours, minutes and seconds?. 

---

.pull-left[
```{r lu02a}
library(lubridate)

ymd(today1)
ymd(today2)
ymd(today3)
```
]
.pull-right[
```{r lu02b}
ymd_hms(today4)
ymd_hms(today5)
mdy_hms(today6)
dmy_hms(today7)
```
]

---

If I need to extract data/time elements, I can do that as well. 

```{r lu07}
today = dmy_hms(today7); today # a date and time set in today7 
today.y = year(today); today.y # The year 
today.m1 = month(today); today.m1 # the month, as a number  
today.m2 = month(today, label = TRUE, abbr = TRUE); today.m2 # labeling the month but with an abbreviation 
today.m3 = month(today, label = TRUE, abbr = FALSE); today.m3# fully labelling the month 
```

---

.pull-left[
```{r lu08a}
today.w = week(today); today.w # what week of the year is it? 
today.doy = yday(today); today.doy # what day of the year is it? 
today.dom = mday(today); today.dom # what day of the month is it? 
```
]
.pull-right[
```{r lu08b}
today.dow1 = wday(today); today.dow1 # what day of the week is it,  as a number? 
today.dow2 = wday(today, label = TRUE, abbr = TRUE); today.dow2 #  day of the week labelled but abbreviated 
today.dow3 = wday(today, label = TRUE, abbr = FALSE); today.dow3 # day of the week fully labelled  
```
]

---

```{r lu08c}
today.h = hour(today); today.h # what hour is it? 
today.m = minute(today); today.m # what minute is it? 
today.s = second(today); today.s # what second is it? 
today.tz = tz(today); today.tz # what time zone is it? 
```

