---
title: "Rmd file for Module 05"
author: "Ani Ruhil"
date: "`r Sys.Date()`"
output:
    html_document: 
      collapsed: no
      highlight: haddock
      smooth_scroll: no
      theme: readable
      toc: yes
      toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', echo = TRUE, warning = FALSE, message = FALSE, dpi = 600, cache = TRUE, tidy = TRUE, tidy.opts = list(width.cutoff = 60), fig.align = "center", fig.width = 10, fig.height = 8, out.width = "65%", dev = 'svg', highlight = TRUE) 
```

# Agenda 
If you want to do anything with dates, it helps to have them formatted correctly, and that is not always the case with secondary data we grab for analysis. Consequently, in this module we will see ways to work with date and time variables via `base R` and `lubridate`, a special package to work with date/time data. We'll start by creating some data values.  

```{r dates00}
basedate = as.Date("1970-1-1")
today = Sys.Date()
tomorrow = as.Date("2017-12-18")
yesterday1 = "12/16/2017"
yesterday2 = "12-16-17"
yesterday3 = "12 16 17"
```

Now let us see how R treats each of these. 

```{r dates01}
str(list(basedate, today, tomorrow, yesterday1, yesterday2, yesterday3))
```

Clearly, only `basedate`, `today` and `tomorrow` are read as date fields; the other three are read as characters. 

## base R 
This latter group can be flipped into date fields: 

```{r dates02}
yesterday1d = as.Date(yesterday1, format = "%m/%d/%Y")
yesterday2d = as.Date(yesterday2, format = "%m-%d-%y")
yesterday3d = as.Date(yesterday3, format = "%m %d %y")
yesterday1d; yesterday2d; yesterday3d
```

There are special switches that indicate date components when you `format = ""`

```{r dates08, echo=FALSE}
df = data.frame(Code = c("%a", "%A", "%d", "%m", "%b", "%B", "%y", "%Y"), 
    Represents = c("Day spelled out (abbreviated)", "Day spelled out", "Day of the month", "Month", "Month (abbreviated)", 
        "Month (fully spelled out)", "Year (2-digits)", "Year (4-digits)"), 
    Example = c("Mon", "Monday", "16", "12", "Dec", "December", "17", "2017"))

library(kableExtra)
knitr::kable(df, booktabs = TRUE, "html") %>% kable_styling(full_width = FALSE)
```

So if you ever run into a date field that needs formatting, what we have covered should help you convert it into a proper date formatted variable.  

Once you convert them to the date format, you can extract other quantities. 

```{r dates03}
weekdays(yesterday1d)
months(today)
quarters(today)
```

Sometimes you will have a number representing the date, with the number representing the number of days since (if positive) or before (if negative) some date  of origin. R uses `1970-01-01` but other software packages may (and some do) use different days. Thus, for example, if the dates are listed as 

```{r dates04}
x = c(17651, -2345, 19760)
```

Then these numbers represent the following dates: 

```{r dates05}
x.dates = as.Date(x, origin = "1970-01-01")
x.dates
```

You can also convert dates into the numbers representing dates.^[These are [Julian days](http://aa.usno.navy.mil/data/docs/JulianDate.php)] 

```{r dates06}
x.julian = julian(x.dates, origin = as.POSIXct("1970-01-01", tz = "UCT"))
x.julian
```

What if I wanted to create a sequence of dates, starting with Feb 8, 2018 but increasing by 1 day at a time or by every 5th day? 

```{r dates07}
start.date = as.Date(as.character("2018-02-08"), format = "%Y-%m-%d")
date.seq1 = seq(start.date, by = 1, length.out = 7)
date.seq1
date.seq5 = seq(start.date, by = 5, length.out = 3)
date.seq5
```

Note that `length.out =` is specifying how many you dates want to create. 


Now, what if I want to know the date 30 days from today? 19 days ago? 

```{r dates09}
date.30 = today + 30
date.19 = today - 19
```

How many days lapsed between two dates? 

```{r dates10}
date1 = as.Date("2017-04-28")
date2 = Sys.Date()
lapsed.time = date2 - date1; lapsed.time
```

Say I want to create a vector of dates that starts and ends on specific dates, and the step function is 1 day, a week, 4 months, etc. The step is indicated with the `by = ""` command 

```{r dates11}
my.dates1 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "day")
my.dates2 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "week")
my.dates3 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "month")
my.dates4 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "3 days")
my.dates5 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "2 weeks")
my.dates6 = seq(from = as.Date("2017-12-17"), to = as.Date("2018-12-16"), by = "4 months")
my.dates7 = seq(from = as.Date("2017-12-17"), to = as.Date("2019-12-16"), by = "year")
my.dates8 = seq(from = as.Date("2017-12-17"), to = as.Date("2022-12-16"), by = "2 years")
```

We can also extract components,  say from dates in the `economics` data-set.

```{r dates12}
library(ggplot2)
data(economics); names(economics)
economics$year = format(economics$date, "%Y")
economics$month = format(economics$date, "%B")
economics$quarter = quarters(economics$date)

head(economics[,  c(1, 7:9)])
```

## lubridate 

This package makes working with dates and times quite easy because it does what base R does but in more intuitive ways and perhaps more flexibly. Let us start with some date fields.

```{r  lu01}
today1 = "20171217"
today2 = "2017-12-17"
today3 = "2017 December 17"
today4 = "20171217143241"
today5 = "2017 December 17 14:32:41"
today6 = "December 17 2017 14:32:41"
today7 = "17-Dec, 2017 14:32:41"
```

The formats are quite varied but `lubridate` deals with them quite seamlessly so long as you pay attention to the order -- is year first or last? What about month? day? Is time given in hours, minutes and seconds?. 

```{r lu02}
library(lubridate)
ymd(today1)
ymd(today2)
ymd(today3)
ymd_hms(today4)
ymd_hms(today5)
mdy_hms(today6)
dmy_hms(today7)
```

The default time zone will  be `UTC`, the Coordinated Universal Time clock. If you want to change it you must specify the time zone.

```{r lu03}
dmy_hms(today7, tz = "EST")
```

I am planning my spring break travel to Italy and want to know when my flight departs on March 10. 

```{r lu04}
now.charleston = now("America/New_York"); now.charleston
my.departure = now.charleston + days(30); my.departure

```

To look-up time zones, run 

```{r lu05}
grep("America",  OlsonNames(), value =  TRUE)
```

Hang on a second, the boarding pass shows departure time to be at 11:41 AM. Ah, easily corrected: 

```{r lu06}
corrected.departure = update(my.departure, hours = 11, minutes = 41, seconds = 0); corrected.departure
```

If I need to extract data/time elements, I can do that as well. 

```{r lu07}
today = dmy_hms(today7); today # a date and time set in today7 
today.y = year(today); today.y # The year 
today.m1 = month(today); today.m1 # the month, as a number  
today.m2 = month(today, label = TRUE, abbr = TRUE); today.m2 # labeling the month but with an abbreviation 
today.m3 = month(today, label = TRUE, abbr = FALSE); today.m3# fully labelling the month 
today.w = week(today); today.w # what week of the year is it? 
today.doy = yday(today); today.doy # what day of the year is it? 
today.dom = mday(today); today.dom # what day of the month is it? 
today.dow1 = wday(today); today.dow1 # what day of the week is it,  as a number? 
today.dow2 = wday(today, label = TRUE, abbr = TRUE); today.dow2 #  day of the week labelled but abbreviated 
today.dow3 = wday(today, label = TRUE, abbr = FALSE); today.dow3 # day of the week fully labelled  
today.h = hour(today); today.h # what hour is it? 
today.m = minute(today); today.m # what minute is it? 
today.s = second(today); today.s # what second is it? 
today.tz = tz(today); today.tz # what time zone is it? 
```

See the book chapter for more advanced work with dates and times and some practice exercises. 

